---
title: "Basic usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  knitr::opts_chunk$set(fig.width=6, fig.height=5), 
  collapse = TRUE,
  comment = "#>"
)
```
## Purpose

This package was born as a collaboration between a physicist and an experimental biologist.
We faced a problem: our department had bought a fancy tracker for registering the positions of macroinvertebrates on a set of Petri dishes.

Long story short, the device captured several images per second, and managed to identify the horizontal and vertical position of the organism under study.
The output produced by the software contained time series with values of the sampling times $t$, and the corresponding two-dimensional positions $(x, y)$.
Unfortunately, this output also contained some obscure and difficult to manage processed data.

We decided to write our own analysis software for these sampled trajectories, using only the fundamental infromation about movement (that is: $t$, $x$ and $y$).
Our packages returns different properties of that movement, such as the instantaneous speeds, accelerations or curvatures.

## Installation

In order to use the package, we have to install it.
The easiest way to install it is via the [`devtools`](https://devtools.r-lib.org/) official CRAN package.

```r
devtools::install_github("PabRod/footprints")
```

After installing, we have to attach the package as usual:

```{r setup}
library(footprints)
```

And we are ready to go!

## Basic usage

### Storing sampled movement

This is a package for studying movement.
So we need a way of coding movement into an object.
This is R, so you guessed it! We'll code movement in the form of a data frame.
Particularly, our data frames representing movement will have three columns:

- The $t$ column: representing time
- The $x$ column: representing the horizontal position at the time $t$
- The $y$ column: representing the vertical position at the time $t$

A minimal example could be:

```{r parabolic-create}
mov <- data.frame(t = c(0, 1, 2, 3, 4), 
                  x = c(0, 1, 2, 3, 4), 
                  y = c(0, 1, 4, 9, 16))
```

that represents five steps of what seems to be a parabolic movement:

```{r parabolic-plot}
plot(mov$x, mov$y, xlab = "x", ylab = "y")
```

Notice that the insides of the data frame look like below:

```{r parabolic-display, echo=FALSE}
mov
```

### Extracting movement properties

The function `append_dynamics` is a way of extracting as much movement properties as possible.
If we apply it to `mov`, the resulting data frame contains much more columns:

```{r append}
mov_analyzed <- append_dynamics(mov)
```

```{r display, echo=FALSE}
mov_analyzed
```

These new columns are:

- `vx`, `vy` and `aspeed`: horizontal, vertical and absolute speed.
- `ax`, `ay`, and `aaccel`: horizontal, vertical and absolute acceleration.
- `curv` and `curv_radius`: curvature and curvature radius.
- `disp_x`, `disp_y` and `adisp`: horizontal, vertical and absolute displacement (since previous time step).

#### A note about units

The units of $x$ and $y$ are expected to be the same (say, meters).
All the new columns derive their units from thse of $t$ and $x$ that is:

- Speeds' units are: $x$ unit divided by $t$ unit.
- Acceleration's units are: $x$ unit divided by $t$ unit squared.
- Curvature unit is $1$ divided by $x$ unit.
- Curvature radius' unit is the same as $x$.
- Displacements' units are the same as $x$.

The new columns are ready to be analyzed.
For example, we can visualize curvature and absolute speed using `ggplot2`:

```{r ggplot}
library(ggplot2)
ggplot(data = mov_analyzed, 
       mapping = aes(x = x, y = y, col = curv_radius, size = aspeed)) +
       geom_point() +
       scale_color_gradient(low="blue", high="red")
```

## More realistic examples

### Parametric equation of movement

A common way of describing movement in physics is by assigning a function of time to $x$ and $y
For instance, an Archimedean spiral is known to follow the equations below:

$$
x(t) = t \cdot \cos(t)
\\
y(t) = t \cdot \sin(t)
$$

We can encode it, and sample at intervals of 0.05 units of time, using the snippet below:

```{r spiral create}
# Generate the times
ts <- seq(0, 10, by = 0.05)

# Calculate the positions using a function
xs <- ts * cos(ts)
ys <- ts * sin(ts)

# Store as data frame
mov <- data.frame(t = ts, 
                  x = xs, 
                  y = ys)
```

The data looks like this:

```{r spiral-plot}
plot(mov$x, mov$y, xlab = "x", ylab = "y", asp = 1)
```

And repeating the analysis shown in the previous example, we can significantly enrich the features we can see on the data.
For instance, once again, we use color to plot curvature radii and size to plot absolute speed.

```{r spiral-analyzed}
mov_analyzed <- append_dynamics(mov)
```

```{r spiral-ggplot}
ggplot(data = mov_analyzed, 
       mapping = aes(x = x, y = y, col = curv_radius, size = aspeed)) +
       geom_point(alpha = 0.5) +
       coord_fixed() +
       scale_color_gradient(low="blue", high="red")
```

### Sampled curve

Much more interesting examples will come from actual **sampled** data, such as the ones that our device produces.
In that case, we'll have to load the data from an external source, and transform it to the (t, x, y) format.
Sometimes, we'll also need to clean the data a bit.

For pedagogical purposes, we include an example data set in this package.

This is how our experimental data looks after importing and cleaning:

```{r load-data}
mov <- footprints::example_mov
plot(mov$x, mov$y, xlab = "x", ylab = "y", asp = 1)

```

And this is an example of how can it look after analyzing:

```{r analyze-data}
mov_analyzed <- append_dynamics(mov)

ggplot(data = mov_analyzed, 
       mapping = aes(x = x, y = y, col = aaccel, size = aspeed)) +
       geom_point(alpha = 0.1) +
       coord_fixed() +
       scale_color_gradient(low="blue", high="red")
```
