---
title: "Example of usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Purpose

This package was born as a collaboration between a physicist and an experimental biologist.
We faced a problem: our department had bought a fancy tracker for registering the positions of macroinvertebrates on a set of Petri dishes.
The device came with its own software, but the output happened to be quite obscure and unmanageable.

We decided to write our own analysis software for these sampled trajectories.
It works with a main input: a data frame representing the sampled movement on a two-dimensional surface.
It returns different properties of that movement, such as the instantaneous speeds, accelerations or curvatures.

## Installation

In order to use the package, we have to install it.
In case you haven't done it already, the easiest way to install it is via the `devtools` official CRAN package.

```r
devtools::install_github('PabRod/footprints')
```

After installing, we have to attach the package.

```{r setup}
library(footprints)
```

And we are ready to go!

## Usage

### Storing sampled movement

This is a package for studying movement.
So we need a way of coding movement into an object.
This is R, so you guessed it! We'll code movement in the form of a data frame.
Particularly, our data frames representing movement will have three columns:

- The t column: representing time
- The x column: representing the horizontal position at the time t
- The y column: representing the vertical position at the time t

An example could be:

```{r}
mov <- data.frame(t = c(0, 1, 2, 3, 4), 
                  x = c(0, 1, 2, 3, 4), 
                  y = c(0, 1, 4, 9, 16))
```

that represents five steps of what seems to be a parabolic movement:

```{r}
plot(mov$x, mov$y, xlab = 'x', ylab = 'y')
```

More interesting example will come from **sampled** data frames, such as the ones that our device produces.
In that case, we'll have to load the data from an external source, and transform it to the (t, x, y) format.

